---
title: "Nuclear Energy in the Global Power Mix"
author: "Akima M. Terrones"
date: "2025-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

Nuclear energy is an important source of low-carbon electricity, playing a role in helping countries meet rising energy needs while reducing greenhouse-gas emissions. However, nuclear energy adoption has not been the worldwide 1st approach to energy. Some countries have expanded nuclear energy production, while others have reduced output or phased out facilities, especially after major events such as its use in weaponry and power plant accidents.

This project examines historical nuclear electricity production across countries to identify long-term trends, compare major producers, and understand how global patterns may reflect policy, safety, and environmental considerations. This analysis aims to provide insight into nuclear energy’s place in today’s energy landscape and its potential role in the future.

## About the Dataset

The dataset used in this project comes from **Our World in Data**, an open-access research platform that compiles global historical information about energy, climate, and development. The file includes country-level observations of nuclear electricity production over time.

Key variables include:

- **country** — Name of the country or region
- **code** — Standard three-letter country code
- **year** — Year of measurement
- **nuclear_TWh** — Electricity produced from nuclear power (terawatt-hours)

The data spans multiple countries and years, allowing us to observe long-term nuclear energy trends globally. It allows comparisons across countries and time. This dataset is useful for exploring how nuclear power has changed, which nations produce the most, and how external events or policy changes may have influenced energy production.

```{r Nuclear data, include=FALSE}
##1 Nuclear Data
library(tidyverse)
library(lubridate)

nuclear <- read_csv("nuclear-energy-generation.csv") %>%    ##comes from website our world in data website for Nuclear Energy
  rename(
    country     = Entity,
    code        = Code,
    year        = Year,
    nuclear_TWh = `Electricity from nuclear - TWh`
  )

glimpse(nuclear)
colnames(nuclear)

global_nuclear <- nuclear %>%
  filter(!is.na(nuclear_TWh)) %>%
  group_by(year) %>%
  summarize(global_gen = sum(nuclear_TWh, na.rm = TRUE))
```

## Questions
1) Which countries have had the highest nuclear energy production over time, and how have their outputs changed?
  planned graph: TBD
2) How has global nuclear electricity generation changed over time?
  planned graph: Line chart of global nuclear_TWh over year
3) How does nuclear energy usage differ by country or region?
  planned graph: Boxplot or grounded bar chart
4) Did nuclear generation trends shift after major historical events such as the 1986 Chernobyl disaster or the 2011 Fukushima accident?
  planned graph: Before/after comparison timeline plot
5) What do these trends suggest about nuclear energy’s role in supporting global decarbonization and energy security?

## Data observations and Visualizations
#### Which countries have had the highest nuclear energy production over time, and how have their outputs changed?

The dataset identifies the highest total nuclear electricity producers as the United States, France, Japan, Russia, and Germany, with an additional aggregated series labeled “World”, which represents global nuclear production as a whole. 

Among these, the United States has consistently led in nuclear electricity generation, maintaining a significant lead over other countries throughout the observed period, reflecting its early investment in nuclear technology and its large reactor fleet. France also shows sustained high output due to it's long-standing national strategy to generate most of its electricity from nuclear power. 

Japan and Germany show noticeable declines after 2011. Japan’s sharp drop corresponds with the shutdown of many reactors following the Fukushima Daiichi accident, while Germany’s decline reflects its commitment to phasing out nuclear energy entirely. Russia displays a steadier long-term trend, maintaining moderate but consistent levels of nuclear production.

```{r question1, echo=FALSE}
library(tidyverse)
# Filter
nuclear_countries <- nuclear %>%
  filter(!is.na(code))   # keep only rows with a 3-letter country code

# Identify top nuclear-producing countries by total generation
top_countries <- nuclear_countries %>%
  group_by(country) %>%
  summarize(total_gen = sum(nuclear_TWh, na.rm = TRUE)) %>%
  slice_max(total_gen, n = 6) %>%   # top 6 producers
  pull(country)
top_countries

# Multi-timeline trend
nuclear_countries %>%
  filter(country %in% top_countries) %>%
  ggplot(aes(x = year, y = nuclear_TWh, color = country)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1, alpha = 0.8) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Nuclear Electricity Production Over Time",
    subtitle = "Top nuclear-producing countries (by total TWh generated)",
    x = "Year",
    y = "Nuclear Electricity Generation (TWh)",
    color = "Country"
  ) + theme_minimal(base_size = 13) + theme( plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(color = "gray40"),panel.grid.minor = element_blank()
  )

```
These country-level trends highlight how nuclear production differs widely based on national policies and historical events. However, to understand the broader impact of these changes, it’s important to look beyond individual countries and examine how global nuclear electricity generation has shifted over time.

#### How has global nuclear electricity generation changed over time?

The global trend in nuclear electricity generation shows sustained growth from the 1970s through the early 2000s. Even after the 1986 Chernobyl disaster, worldwide nuclear output continued to increase. This reflects the fact that most countries did not immediately halt or reduce nuclear production, and many reactors built in earlier decades were still expanding or coming online in the years following the accident.
```{r question2, echo=FALSE, message=FALSE, warning=FALSE}
library(dygraphs)
library(xts)

global_xts <- xts(global_nuclear$global_gen,
                  order.by = as.Date(paste0(global_nuclear$year, "-01-01")))

dygraph(global_xts, main = "Global Nuclear Energy Generation Over Time") %>%
dySeries("V1", label = "TWh", color = "green") %>%
  dyRangeSelector() %>%
  dyOptions(fillGraph = TRUE, fillAlpha = 0.3, colors = "green")

```
A noticeable shift emerges after the 2011 Fukushima accident. In the years immediately following Fukushima, global nuclear generation declines as countries such as Japan, Germany, and others shut down reactors or tightened regulatory standards. This suggests that while nuclear energy remained a significant part of the global energy mix, concerns about safety and public opposition following major accidents influenced the pace of new reactor construction and led some countries to reduce their reliance on nuclear power. However, this decline is not permanent.

Shortly after 2012, nuclear energy generation increases, indicating that it was still being used as a major source of low-carbon electricity. This rebound may reflect ongoing investments in nuclear technology, the commissioning of new reactors within countries and the recognition of nuclear power’s role in meeting climate goals. 

Soon after some time, in 2023 we witness a steep ongoing decline that could be attributed to various reasons such a combination of aging reactor retirements, slower construction timelines for new facilities, and evolving national energy policies as the population increases and energy demands grow.

Overall, the data show that while major nuclear accidents can influence public perception and policy decisions, nuclear energy remains a resilient and substantial contributor to the world’s electricity supply. Its long-term stability highlights its continued role as a reliable, large-scale, low-emission energy source within the global mix.

#### How does nuclear energy usage differ by country or region?

The distribution of nuclear electricity generation varies sharply across global regions, revealing clear structural differences in how countries rely on nuclear power. North America stands out with the highest and most widely spread levels of nuclear production. This is largely driven by the United States, which operates the world’s largest fleet of nuclear reactors and consistently generates more nuclear electricity than any other nation.

Europe shows moderately high but less variable output, reflecting the long-term stability of nuclear programs in countries like France, Germany (before its phase-out), and Sweden. Europe’s narrower spread suggests that production levels have been more consistent over time, even as individual nations adjusted their energy policies.

Asia displays a wider and more uneven distribution, capturing the diverse nuclear trajectories of countries such as Japan, South Korea, China, and India. China’s rapid nuclear expansion contrasts with Japan’s post-Fukushima decline, creating a broad range of production values within the region.

In contrast, Latin America and the Other category maintain very low levels of nuclear output. While these regions include countries with emerging or limited nuclear infrastructure, their consistently small distributions highlight the concentration of nuclear technology in a few major regions.

```{r question 3, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)

# Create simple regions using case_when
nuclear_region <- nuclear %>%
  filter(!is.na(code)) %>%
  mutate(region = case_when(
    country %in% c("United States", "Canada") ~ "North America",
    country %in% c("France", "Germany", "United Kingdom", "Sweden", "Spain", "Belgium") ~ "Europe",
    country %in% c("Japan", "South Korea", "China", "India") ~ "Asia",
    country %in% c("Brazil", "Argentina", "Mexico") ~ "Latin America",
    TRUE ~ "Other"
  ))

# Filter years
nuclear_region_filtered <- nuclear_region %>% 
  filter(year >= 1990)

# ggplot
p <- nuclear_region_filtered %>%
  ggplot(aes(x = region, y = nuclear_TWh, fill = region)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Nuclear Electricity Generation by Region",
    subtitle = "Distribution of annual nuclear electricity production (TWh), 1990–2024",
    x = "Region",
    y = "Nuclear Electricity Generation (TWh)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    legend.position = "none"
  )

# Convert to interactive plotly
ggplotly(p)

```
Overall, the regional comparison shows that nuclear energy remains geographically concentrated, with North America, Europe, and parts of Asia producing the vast majority of global nuclear electricity. These differences reflect regional policy choices, technological capacity, historical investment, and the broader role nuclear energy plays in each region’s energy strategy.

#### Did nuclear generation trends shift after major historical events such as the 1986 Chernobyl disaster or the 2011 Fukushima accident?

In this chart, the two portions highlighted are two major nuclear accident, Chernobyl (1986) and Fukushima (2011) -as shown in graph 2 as well- This visualization shows the global response over the years that followed the events.

Following Chernobyl, global output did not decline; instead, it continued rising through the late 1980s and 1990s, suggesting that most countries maintained or expanded their nuclear programs despite heightened concern.

Fukushima, however, shows a more immediate global impact. After 2011, nuclear generation plateaus and then gradually declines as countries like Germany and Japan reduced or shut down reactors and global safety standards tightened. This pattern indicates that Fukushima produced a stronger and more sustained shift in worldwide nuclear policy than Chernobyl.

```{r question4, echo=FALSE, message=FALSE, warning=FALSE}
# Get max for placing labels nicely
max_y <- max(global_nuclear$global_gen, na.rm = TRUE)

global_nuclear %>%
  ggplot(aes(x = year, y = global_gen)) +
  geom_line(linewidth = 1.1, color = "#1b9e77") +
  
  # Shaded windows around major accidents
  annotate("rect", xmin = 1986, xmax = 1990,
           ymin = -Inf, ymax = Inf,
           fill = "red", alpha = 0.06) +
  annotate("rect", xmin = 2011, xmax = 2015,
           ymin = -Inf, ymax = Inf,
           fill = "steelblue", alpha = 0.06) +
  
  # Vertical dashed lines at event years
  geom_vline(xintercept = 1986, linetype = "dashed", color = "red3") +
  geom_vline(xintercept = 2011, linetype = "dashed", color = "steelblue4") +
  
  # Text labels for the events
  annotate("text", x = 1986, y = max_y * 0.95,
           label = "Chernobyl (1986)", angle = 90,
           vjust = -0.3, size = 3) +
  annotate("text", x = 2011, y = max_y * 0.95,
           label = "Fukushima (2011)", angle = 90,
           vjust = -0.3, size = 3) +
  
  labs(
    title = "Global Nuclear Electricity Generation with Major Accidents Highlighted",
    subtitle = "Total global nuclear generation (TWh) with Chernobyl (1986) and Fukushima (2011) marked",
    x = "Year",
    y = "Global Nuclear Electricity Generation (TWh)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )
```

Overall, the graph suggests that Chernobyl triggered long-term regulatory change but did not halt global nuclear growth, whereas Fukushima produced a noticeable pause and soft decline in worldwide nuclear output, reflecting a stronger global policy response and heightened public concern.

#### What do these trends suggest about nuclear energy’s role in supporting global decarbonization and energy security?

In my graph, what is is highlighted is how nuclear energy's contribution to electricity generation is distributed across global regions in the modern era. North America and Europe clearly dominate nuclear output, producing substantially higher average annual generation compared to other regions. This reflects decades of established reactor capacity, consistent policy support, and investments in long-term energy security.

Regions such as Asia display meaningful but more moderate levels of nuclear generation, suggesting ongoing expansion but not yet at the scale of the largest producers. Meanwhile, Latin America, the Middle East, Africa, and other regions show minimal nuclear output, indicating limited infrastructure, higher financial barriers, or stronger reliance on alternative energy sources.

```{r question 5, echo=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)

# Average nuclear generation by region in recent decades
q5_summary <- nuclear_region %>%
  filter(year >= 2000, nuclear_TWh > 0) %>%      #modern period
  group_by(region) %>%
  summarize(avg_TWh = mean(nuclear_TWh, na.rm = TRUE)) %>%
  arrange(desc(avg_TWh))

# Static bar plot
p5 <- q5_summary %>%
  ggplot(aes(x = reorder(region, avg_TWh),
             y = avg_TWh,
             fill = region)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Nuclear Electricity Generation by Region",
    subtitle = "Mean annual nuclear generation (TWh), 2000–2024",
    x = "Region",
    y = "Average Nuclear Electricity Generation (TWh)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "Black"),
    legend.position = "none"
  )

ggplotly(p5)
```

Together, these patterns suggest that nuclear energy plays a central role in decarbonization and energy security primarily in regions that have already built large, stable nuclear fleets. In these regions, nuclear power contributes reliable, low-carbon baseload electricity.

Overall, the evidence suggests that while nuclear energy is not without risks, it remains a critical component of the global clean-energy mix, offering reliable baseload power that complements intermittent renewable sources such as solar and wind. Its long-term stability and low emissions position it as an important pillar for meeting climate goals while ensuring consistent and secure access to electricity.

## Conclusion

In conclusion, the data shows that nuclear energy has remained a steady and important source of low-carbon electricity around the world. The top-producing countries continue to generate large amounts of nuclear power, and global trends highlight long periods of growth with only temporary slowdowns after major events. Even when accidents like Fukushima created policy shifts, nuclear energy still remained a major part of the global energy mix.

Regionally, North America, Europe, and parts of Asia carry most of the world’s nuclear production, while other regions contribute far less. This uneven distribution reflects differences in infrastructure, investment, and national priorities.

Taken together, these trends suggest that nuclear energy plays a strong role in supporting decarbonization and energy security. It provides reliable, large-scale, low-carbon electricity that helps countries meet energy needs while reducing emissions. Even with challenges, nuclear power continues to be a key part of the global clean-energy landscape.
